Preparing data from the Global Burden of Disease 2019 for METAHIT modelling

### 2022-07-12 metahit/paper_analyses.Rmd
Prepare for CRAN release

ggplot(gbdplot, aes(x=age, y=mort_prob, col=area, lty=gender, group=areagender)) + 
  facet_wrap(~disease, ncol=3, scales="free_y") + 
  coord_cartesian(xlim = c(50,100)) +
  ylab("Annual mortality probability") + 
  xlab("Year of age") +
  geom_ribbon(aes(ymin=mort_lower, ymax=mort_upper), fill="gray80", col="gray80") +
  geom_line() + 
  labs(col = "City region") +
  theme_bw() + 
  theme(strip.text.x = element_text(size = 7.5),
        legend.position = "none")
dev.off()

pdf("../../write/inc_data.pdf", height=ht)
ggplot(gbdplot, aes(x=age, y=inc_prob, col=area, lty=gender, group=areagender)) + 
  facet_wrap(~disease, ncol=3, scales="free_y") + 
  coord_cartesian(xlim = c(50,100)) +
  ylab("Annual incidence probability") + 
  xlab("Year of age") +
  geom_ribbon(aes(ymin=inc_lower, ymax=inc_upper), fill="gray80", col="gray80") +
  geom_line() + 
  labs(col = "City region") +
  theme_bw() + 
  theme(strip.text.x = element_text(size = 7.5),
        legend.position = "none")
dev.off()

@@

resopt <- readRDS("resopt_nonhier.rds") %>% 
  filter(!disease %in% diseases_nat)
resoptnat <- readRDS("resopt_nat.rds")
resoptcf <- resopt %>%
  full_join(resoptnat) %>% 
  filter(var=="cf") %>%
  mutate(p50 = 1 - exp(-`mode`),
         plow = 1 - exp(-`2.5%`),
         pupp = 1 - exp(-`97.5%`)) %>%
  filter(area=="Bristol" | area=="England", gender=="Female") %>%
  mutate(disease = fct_recode(disease, !!!disease_shorten),
           disease = factor(disease, levels = disease_order)) %>%
  filter(age %in% c(70, 75, 80, 85, 90)) %>%
  mutate(Algorithm = "Mode approximation")

@@

resp <- resall %>% 
  dplyr::filter(model %in% c("Independent areas","National"), var=="cf") %>%
  dplyr::filter((area=="Bristol" & (!(disease %in% diseases_nat))) |
                    (area=="England" & model=="National"), 
                gender=="Female") %>%
  mutate(Algorithm="MCMC") %>%
  full_join(resoptcf)
resp %>%
  filter(disease %in% diseases_paper, 
         age %in% c(70, 75, 80, 85, 90)) %>%
  mutate(par = 1 - exp(-`50%`),
         plow = 1 - exp(-`2.5%`),
         pupp = 1 - exp(-`97.5%`)) %>%
  ggplot(aes(x=age, y=par, col=Algorithm, lty=Algorithm)) + 
  geom_errorbar(aes(ymin=plow, ymax=pupp), width=1.5,
              position=position_dodge(width=2), lwd=1.3) +
  geom_point(position=position_dodge(width=2), size=1.3) + 
  facet_wrap(vars(disease), ncol=2, scales="free_y") + 
  xlab("Age (years)") + ylab("Case fatality probability")

@@

resplotnh <- resnh %>% 
    dplyr::filter(model %in% c("Independent areas","National"), 
                  var=="cf",
                area %in% c(cityregions,"England")) %>%
    mutate(area = forcats::fct_relevel(area, "England", after=Inf)) %>%
    dplyr::filter(! (disease=="Uterine cancer" & age > 90),
                  disease %in% diseases_paper) %>%
    mutate(p50 = 1 - exp(-`50%`),
         plow = 1 - exp(-`2.5%`),
         pupp = 1 - exp(-`97.5%`))

gg <- resplotnh %>%
  ggplot(aes(x=age, y=p50, group=area, col=area)) + 
  geom_ribbon(aes(ymin=plow, ymax=pupp), 
              fill="gray80", col="gray80") +
  geom_line() + 
  facet_grid(cols=vars(gender), rows=vars(disease), scales = "free_y", switch = "y") + 
  xlab("Age (years)") + 
  ylab("Annual case fatality probability") + 
  xlim(60,100) + 
  theme(strip.text.y.left = element_text(angle = 0),
        strip.text.y = element_text(size=6),
        axis.text = element_text(size=8),
        legend.text = element_text(size=6)) + 
    labs(col="")

@@

## UTRC doesn't converge, and STMC estimates noisy, so we use national estimate results 
resnc <- resh %>% filter(Rhat > 2) 
resnc <- resh %>% filter(Rhat > 1.1) 
resnc <- resh %>% filter(Rhat > 1.05) 
table(resnc$disease, resnc$gender)

@@

## Between area variability in age curves, point ests.  All diseases 
rescfplot <- resall %>% 
    filter(var=="cf", 
           disease %in% diseases_paper_short, 
           area %in% c(cityregions, "England"),
           (model %in% c("Independent areas","National")) | 
               (model %in% "Hierarchical" & !(disease %in% diseases_nat))
    ) %>%
        mutate(disease = fct_recode(disease, 
                                    "Colorectal cancer"="Colon and rectum cancer",
                                    "IHD" = "Ischemic heart disease")) %>%
    mutate(model = ifelse(model=="National","Independent areas",model),
           p50 = 1 - exp(-`50%`),
           plow = 1 - exp(-`2.5%`),
           pupp = 1 - exp(-`97.5%`),
           gender_model = paste(gender, model, sep="\n")) %>%
    mutate(gender_model = factor(gender_model, 
                                 levels=c("Female\nIndependent areas",
                                          "Male\nIndependent areas",
                                          "Female\nHierarchical",
                                          "Male\nHierarchical"))) 
gg <- rescfplot %>%
    ggplot(aes(x=age, col=area, group=area)) +
    geom_ribbon(aes(ymin=plow, ymax=pupp), fill="gray80", col="gray80") +
    geom_line(aes(y=p50)) + 
    coord_cartesian(xlim=c(70,  100), ylim=c(0,NA)) +
    scale_x_continuous(breaks=seq(0,100,10)) + 
    ylab("Annual case fatality probability")  +
    facet_grid(cols=vars(gender_model), rows=vars(disease),
               scales="free_y", switch = "y") + 
    xlab("Age (years)") + 
    labs(col="") +
    theme(strip.text.y.left = element_text(angle = 0),
          strip.text.x = element_text(size = 5),
          axis.text.x = element_text(size = 5.5),
          axis.text.y = element_text(size = 5),
          axis.title.x = element_text(size = 7),
          axis.title.y = element_text(size = 7),
          strip.text.y = element_text(size = 6.5),
          legend.text = element_text(size = 6.5))
gg

@@

>>>>

ggplot(gbdplot, aes(x=age, y=mort_prob, col=area, lty=gender, group=areagender)) + 
  facet_wrap(~disease, ncol=3, scales="free_y") + 
  coord_cartesian(xlim = c(50,100)) +
  ylab("Annual mortality risk") + 
  xlab("Year of age") +
  geom_ribbon(aes(ymin=mort_lower, ymax=mort_upper), fill="gray80", col="gray80") +
  geom_line() + 
  labs(col = "City region") +
  theme_bw() + 
  theme(strip.text.x = element_text(size = 7.5),
        legend.position = "none")
dev.off()

pdf("../../write/inc_data.pdf", height=ht)
ggplot(gbdplot, aes(x=age, y=inc_prob, col=area, lty=gender, group=areagender)) + 
  facet_wrap(~disease, ncol=3, scales="free_y") + 
  coord_cartesian(xlim = c(50,100)) +
  ylab("Annual incidence risk") + 
  xlab("Year of age") +
  geom_ribbon(aes(ymin=inc_lower, ymax=inc_upper), fill="gray80", col="gray80") +
  geom_line() + 
  labs(col = "City region") +
  theme_bw() + 
  theme(strip.text.x = element_text(size = 7.5),
        legend.position = "none")
dev.off()

@@

resopt <- readRDS("resopt_nonhier.rds") %>% 
  filter(!disease %in% diseases_nat)
resoptnat <- readRDS("resopt_nat.rds")
resoptcf <- resopt %>%
  full_join(resoptnat) %>% 
  filter(var=="cf") %>%
  mutate(p50 = `mode`,
         plow = `2.5%`,
         pupp = `97.5%`) %>%
  filter(area=="Bristol" | area=="England", gender=="Female") %>%
  mutate(disease = fct_recode(disease, !!!disease_shorten),
           disease = factor(disease, levels = disease_order)) %>%
  filter(age %in% c(70, 75, 80, 85, 90)) %>%
  mutate(Algorithm = "Mode approximation")

@@

resp <- resall %>% 
  dplyr::filter(model %in% c("Independent areas","National"), var=="cf") %>%
  dplyr::filter((area=="Bristol" & (!(disease %in% diseases_nat))) |
                    (area=="England" & model=="National"), 
                gender=="Female") %>%
  mutate(Algorithm="MCMC") %>%
  full_join(resoptcf)
resp %>%
  filter(disease %in% diseases_paper, 
         age %in% c(70, 75, 80, 85, 90)) %>%
  mutate(par = `50%`,
         plow = `2.5%`,
         pupp = `97.5%`) %>%
  ggplot(aes(x=age, y=par, col=Algorithm, lty=Algorithm)) + 
  geom_errorbar(aes(ymin=plow, ymax=pupp), width=1.5,
              position=position_dodge(width=2), lwd=1.3) +
  geom_point(position=position_dodge(width=2), size=1.3) + 
  facet_wrap(vars(disease), ncol=2, scales="free_y") + 
  xlab("Age (years)") + ylab("Case fatality rate")

@@

resplotnh <- resnh %>% 
    dplyr::filter(model %in% c("Independent areas","National"), 
                  var=="cf_prob",
                area %in% c(cityregions,"England")) %>%
    mutate(area = forcats::fct_relevel(area, "England", after=Inf)) %>%
    dplyr::filter(! (disease=="Uterine cancer" & age > 90),
                  disease %in% diseases_paper) 
gg <- resplotnh %>%
  ggplot(aes(x=age, y=`50%`, group=area, col=area)) + 
  geom_ribbon(aes(ymin=`2.5%`, ymax=`97.5%`), 
              fill="gray80", col="gray80") +
  geom_line() + 
  facet_grid(cols=vars(gender), rows=vars(disease), scales = "free_y", switch = "y") + 
  xlab("Age (years)") + 
  ylab("Annual case fatality probability") + 
  xlim(60,100) + 
  theme(strip.text.y.left = element_text(angle = 0),
        strip.text.y = element_text(size=6),
        axis.text = element_text(size=8),
        legend.text = element_text(size=6)) + 
    labs(col="")

@@

## UTRC doesn't converge, and STMC estimates noisy, so we use national estimate results 
## Dementia is only other one where sampling performance is poor.
## Trace plots for dementia look like convergence to plausible region, but poor mixing
resnc <- resh %>% filter(Rhat > 2) 
resnc <- resh %>% filter(Rhat > 1.1, age>80, var=="cf") 
resnc <- resh %>% filter(Rhat > 1.05) 
table(resnc$disease, resnc$gender)

@@

## Between area variability in age curves, point ests.  All diseases 
rescfplot <- resall %>% 
    filter(var=="cf_prob", 
           disease %in% diseases_paper_short, 
           area %in% c(cityregions, "England"),
           (model %in% c("Independent areas","National")) | 
               (model %in% "Hierarchical" & !(disease %in% diseases_nat))
    ) %>%
        mutate(disease = fct_recode(disease, 
                                    "Colorectal cancer"="Colon and rectum cancer",
                                    "IHD" = "Ischemic heart disease")) %>%
    mutate(model = ifelse(model=="National","Independent areas",model),
           gender_model = paste(gender, model, sep="\n")) %>%
    mutate(gender_model = factor(gender_model, 
                                 levels=c("Female\nIndependent areas",
                                          "Male\nIndependent areas",
                                          "Female\nHierarchical",
                                          "Male\nHierarchical"))) 
gg <- rescfplot %>%
    ggplot(aes(x=age, col=area, group=area)) +
    geom_ribbon(aes(ymin=`2.5%`, ymax=`97.5%`), fill="gray80", col="gray80") +
    geom_line(aes(y=`50%`)) + 
    coord_cartesian(xlim=c(70,  100), ylim=c(0,NA)) +
    scale_x_continuous(breaks=seq(0,100,10)) + 
    ylab("Annual case fatality risk")  +
    facet_grid(cols=vars(gender_model), rows=vars(disease),
               scales="free_y", switch = "y") + 
    xlab("Age (years)") + 
    labs(col="") +
    theme(strip.text.y.left = element_text(angle = 0),
          strip.text.x = element_text(size = 5),
          axis.text.x = element_text(size = 5.5),
          axis.text.y = element_text(size = 5),
          axis.title.x = element_text(size = 7),
          axis.title.y = element_text(size = 7),
          strip.text.y = element_text(size = 6.5),
          legend.text = element_text(size = 6.5))
gg

@@

# Discrete-time approximate estimate of case fatality.

Mortality divided by prevalence.

Leeds, female, all diseases.  Use standard disbayes model.

```{r}
setwd("metahit")
source("paper_analyses_header.r")

mhi <- gbd %>%
  mutate(disease = fct_recode(disease, !!!disease_shorten)) %>%
  filter(disease %in% diseases_paper_short) %>%
  filter(area=="Leeds", gender=="Female") %>%
  droplevels() %>%
  mutate(mort_prob = mort_num / mort_denom, 
         prev_prob = prev_num / prev_denom,
         mort_rate = -log(1 - mort_prob),  
         cf_crude = mort_rate / prev_prob)

db <- resall %>% 
  filter(model=="Independent areas") %>%
  filter(area=="Leeds", gender=="Female") %>%
  filter(disease %in% diseases_paper_short) %>%
  filter(var=="cf") %>%
  left_join(mhi %>% select(age, gender, disease, cf_crude))

p1 <- ggplot(db %>% filter(age>60, age <80), 
             aes(x=age, y=`50%`)) + 
  facet_grid(rows=vars(disease), scales = "free_y", labeller = label_wrap_gen(15)) +
  geom_line(lwd=1.5) + 
  geom_ribbon(aes(ymin=`2.5%`, ymax=`97.5%`), alpha=0.3) + 
  geom_line(aes(y=cf_crude), col="red", lwd=1.5, alpha=0.6) + 
  xlab("Age (years)") + 
  ylab("Case fatality rate") +
  theme(strip.text = element_text(size=7))
p1

p2 <- ggplot(db %>% filter(age>=80, age <100), 
             aes(x=age, y=`50%`)) + 
  facet_grid(rows=vars(disease), scales = "free_y", labeller = label_wrap_gen(15)) +
  geom_line(lwd=1.5) + 
  geom_ribbon(aes(ymin=`2.5%`, ymax=`97.5%`), alpha=0.3) + 
  geom_line(aes(y=cf_crude), col="red", lwd=1.5, alpha=0.6) + 
  xlab("Age (years)") + 
  ylab("Case fatality rate") +
  theme(strip.text = element_text(size=7))
p2

pdf("~/work/chronic/write/cf_discrete.pdf")
gridExtra::grid.arrange(p1, p2, nrow=1)
dev.off()

Estimate is non-smooth, unreliable for older ages, doesn't account for uncertainty.

# Comparison of MCMC and optimisation

Timings.

```{r}
system.time(db <- disbayes(data=mhi,
               inc_num = "inc_num", inc_denom = "inc_denom",
               mort_num = "mort_num", mort_denom = "mort_denom",
               prev_num = "prev_num", prev_denom = "prev_denom",
               eqage = 40))
options(mc.cores = 4)
system.time(db <- disbayes(data=mhi,
               inc_num = "inc_num", inc_denom = "inc_denom",
               mort_num = "mort_num", mort_denom = "mort_denom",
               prev_num = "prev_num", prev_denom = "prev_denom",
               eqage = 40, method="mcmc",iter=1000))
```

Compare point estimates and uncertainty, and how these relate to the amount of data, and age.

```{r}
source("read_model_results.r")
rescomp <- resall %>%
  filter(model=="Independent areas") %>%
  mutate(method="MCMC",
         pointest = `50%`) %>%
  full_join(readRDS("resopt_nonhier.rds") %>% 
              mutate(method="Optimisation",
                     pointest = `50%`)) %>%
  filter(
         disease %in% diseases_paper,
         !(disease %in% c("Stomach cancer", "Uterine cancer"))
         ) %>%
  mutate(iqr = `75%` - `25%`,
         ciw = `97.5%` - `2.5%`) %>%
  filter(var == "cf",
         age > 50) %>%
  select(age, gender, disease, area, method, pointest, iqr, ciw) %>%
  left_join(gbd %>% select(age, gender, disease, area, mort_num), 
            by=c("age","gender","disease","area"))

resc_bias <- rescomp %>% 
  select(-iqr, -ciw) %>%
  pivot_wider(names_from=method, 
              values_from=pointest, 
              id_cols = c("age","gender","disease","area","mort_num")) %>%
  mutate(bias = (Optimisation - MCMC),
         dga = interaction(disease, gender, area))

resc_uncbias <- rescomp %>% 
  select(-pointest, -iqr) %>%
  pivot_wider(names_from=method, 
              values_from=ciw, 
              id_cols = c("age","gender","disease","area","mort_num")) %>%
  mutate(bias = 100*(Optimisation - MCMC)/MCMC, 
         dga = interaction(disease, gender, area))

library(gridExtra)
library(ggpubr)

p1 <- ggplot(data=resc_bias, aes(x=age, y=bias, group=dga, col=disease)) + 
  geom_line(alpha=0.5) +
  geom_point() +
  ylab("Median(optimisation) - median(MCMC)") + 
  xlab("Age (years)")

p2 <- ggplot(data=resc_uncbias, aes(x=age, y=bias, group=dga, col=disease)) + 
  geom_line(alpha=0.3) +
  geom_point(alpha=0.5) +
  ylab("%Bias in CI width") +
  xlab("Age (years)")

p3 <- ggplot(data=resc_bias, aes(x=mort_num, y=bias, group=dga, col=disease)) + 
  geom_line(alpha=0.3) +
  geom_point(alpha=0.5) +
  ylab("Median(optimisation) - median(MCMC)") +
  xlab("Number of deaths in the subgroup") +
  theme(legend.title = element_blank(), 
        legend.text = element_text(size=8),
        legend.position = "bottom") + 
  guides(col=guide_legend(nrow=3,byrow=TRUE))

p4 <- ggplot(data=resc_uncbias, aes(x=mort_num, y=bias, group=dga, col=disease)) + 
  geom_line(alpha=0.3) +
  geom_point(alpha=0.5) +
  ylab("%Bias in CI width") + 
  xlab("Number of deaths in the subgroup")

p <- ggarrange(p3, p1, p4, p2, common.legend=TRUE, legend="bottom") # assign to object to work around blank page bug
pdf("~/work/chronic/write/opt_compare_full.pdf")
p
dev.off()
